6.867 Homework 1: Gradient Descent and Regression Methods

### 2016-09-30 paper2/paper.tex
small stuff
done

\documentclass[10pt]{paper}
\usepackage{bm,amsmath,graphicx,chngcntr,subfigure}
\usepackage[margin=0.5in]{geometry}
\pagestyle{empty}
\counterwithin{figure}{section}
\pagenumbering{gobble}
\title{6.857 Homework 1: Gradient Descent and Regressions}

@@

\maketitle
\section{Introduction}

In this paper, we discuss various methods of performing regressions. First, we present the stochastic gradient descent and batch gradient descent algorithms, and discuss the benefits and drawbacks of each. Next, we use gradient descent to perform a linear regression on functions with polynomial and sinusoidal bases. We then improve upon this model with ridge regression, which uses an extra weighting term to prevent overfitting. Finally, we present LASSO estimation, which optimizes this model for the case of sparse weights.

@@

\begin{figure}[ht!]
  \label{fig:gd}
  \centering
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/quadratic_steps_vs_mag}
    % \caption{As the step size increases, the algorithm converges quicker.}
  \end{subfigure}
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
    % \caption{As $\lambda$ increases and $\delta$ decreases, error decreases.}
  \end{subfigure}
   \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
    % \caption{As $\mathbf{x}_0$ approaches $(10, 10)$, the rate of convergence increases.}
  \end{subfigure}
  \caption{The effects of $\lambda$ (left), $\mathbf{x}_0$ (middle), and $\delta$ (right) on gradient descent.}
\end{figure}

@@

With large datasets, it might be inefficient to compute a gradient over the entire dataset for each iteration. One way to address this is to use stochastic gradient descent, which uses a pointwise gradient instead. In this experiment, we use least-squares regression, which has a pointwise gradient of 
\begin{align*}
J(\theta; x^{(i)}, y^{(i)}) = (x^{(i)}\theta_t - y^{(i)})^2
\end{align*}
On each iteration, we update our value as follows:
\begin{align*}
\theta_{t+1} = \theta_{t} - \nu_t \nabla_\theta J(\theta_t; x^{(i)}, y^{(i)})
\end{align*}
where $\nabla = (\tau_0 + t)^{-\kappa}$ is the \emph{learning rate}, and $\tau_0$ and $\kappa$ are constants.

@@

When the objective function doesn't have a closed form solution, one can approximate the gradient using a \emph{central finite difference}. In other words, we have 
\begin{align*}
\nabla f(x) \approx \nabla_h[f](x) = \dfrac{f(x+h/2) - f(x-h/2)}{h}
\end{align*}
Therefore, we only need the objective function to approximate the gradient. Note that as $h$ approaches zero, this approaches the actual value of the gradient; and as $h$ increaes, the value diverges. Figure \ref{fig:cfd}illustrates the impact of $h$ on the accuracy of the gradient: even for large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.

@@

One approach to this problem is to model $f$ as the linear combination of \emph{basis functions}. In other words, we manually choose functions $\phi_1(x), \dots, \phi_n(x)$ and algorithmically pick coefficients $c_1, \dots, c_n$, which we put together as
\begin{align*}bm
f(x) &= c_1 \phi_1(x) + \dots + c_n \phi_n(x).
\end{align*}
We can rewrite this with 
\begin{align*}
\bm{c} &= \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix} \\
\bm{\phi}(x) &= \begin{bmatrix} \phi_1(x) & \cdots & \phi_n(x) \end{bmatrix}
\end{align*}
so that 
\begin{align}
f(x) &= \bm{c} \cdot \bm{\phi}(x). \label{eq:predictor}
\end{align}

@@

To counteract this effect, the learning rate for the stochastic gradient descent needs to be as low as possible for. In this case, we chose $\nu_t = (r + 0.1)^{-0.51}$; with parameters any larger, the result has very little correlation with the actual function. This is likely because a large learning rate would cause smaller-degree terms to have an relatively large weight during the first few iterations, which would overtake the higher-degree terms. Having a smaller, slower-growing learning rate lets the higher-degree terms affect the result for more iterations, and therefore the result more closely fits the input data.

@@

To illustrate in an example, we have three sets of data: set A, set B, and a validation set. We perform the above process twice, once training on A and testing on B, and once vice versa. In both cases we try a variety of values of $\lambda$ and $n$ then evaluate the SSE on the validation set. The results are given in Figure \ref{fig:validationtable}
\begin{figure}[ht!]
  \centering
  \label{fig:validationtable}
  \begin{tabular}{c|cccccc}
    & 1e-05 & 0.0001 & 0.001 & 0.01 & 0.1 \\ \hline
  0 & 62.00 & 62.00 & 62.00 & 62.01 & 62.06 \\
  1 & 2.90 & 2.90 & 2.90 & 2.90 & 2.94 \\
  2 & 2.35 & \textbf{2.35} & 2.35 & 2.36 & 2.43 \\
  3 & 2.37 & 2.37 & 2.37 & 2.39 & 2.55 \\
  4 & 3.58 & 3.58 & 3.58 & 3.59 & 3.67 \\
  5 & 4.14 & 4.14 & 4.14 & 4.12 & 3.97 \\
  6 & 3.98 & 3.98 & 3.98 & 3.96 & 3.89 \\
  7 & 4.07 & 4.07 & 4.06 & 4.03 & 3.89 \\
  \end{tabular}
  \qquad
  \begin{tabular}{c|ccccc}
    & 0 & 0.3 & 0.6 & 0.9 & 1.2 \\ \hline
  0 & 72.44 & 70.95 & 69.63 & 68.47 & 67.44 \\
  1 & 35.19 & 34.09 & 33.14 & 32.34 & 31.67 \\
  2 & 38.53 & 38.65 & 38.76 & 38.87 & 38.96 \\
  3 & 30.30 & 28.46 & \textbf{28.09} & 28.47 & 29.21 \\
  4 & 130.20 & 67.54 & 51.63 & 46.17 & 44.35 \\
  5 & 625.43 & 368.40 & 287.26 & 242.74 & 212.14 \\
  6 & 680.05 & 533.60 & 481.26 & 440.20 & 406.68 \\
  7 & 492.85 & 360.39 & 448.42 & 544.91 & 634.85 \\
  \end{tabular}
  \caption{SSE values on validation set after training with various $n$ (rows) and $\lambda$ (columns). Left: training on set A. Right: training on set B. Best values bolded.}
\end{figure}
The SSE for the first case with $n = 2$ and $\lambda = 0.0001$, tested on set B, is $25.752$. In the second case, tested on set A with $n = 3$ and $\lambda = 0.6$, the SSE is $40.087483697$. Testing with parameters $n=4$ and $\lambda = 0.1$ for both cases, which the table would predict to be suboptimal, indeed gives bigger test SSEs of 31.1388659319 and 64.9392925156, respectively.

@@

>>>>

\documentclass[10pt]{paper}
\usepackage{bm,amsmath,graphicx,chngcntr,subfigure,float}
\usepackage[margin=0.43in]{geometry}
\pagestyle{empty}
\counterwithin{figure}{section}
\pagenumbering{gobble}
\title{6.867 Homework 1: Gradient Descent and Regression Methods}

@@

\maketitle
\vspace{-50px}

\section{Introduction}

In this paper, we discuss various methods of performing regressions. First, we present the stochastic gradient descent and batch gradient descent algorithms, and discuss the benefits and drawbacks of each. Next, we use gradient descent to perform linear regressions. We then improve upon this model with ridge regression, which uses an extra regularization term to prevent overfitting. Finally, we present LASSO estimation, which optimizes this model to encourage sparse coefficient vectors.

@@

\begin{figure}[ht!]
  \label{fig:gd}
  \centering
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/quadratic_steps_vs_mag}
    % \caption{As the step size increases, the algorithm converges quicker.}
  \end{subfigure}
  \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
    % \caption{As $\lambda$ increases and $\delta$ decreases, error decreases.}
  \end{subfigure}
   \begin{subfigure}
    \centering
      \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
    % \caption{As $\mathbf{x}_0$ approaches $(10, 10)$, the rate of convergence increases.}
  \end{subfigure}
  \caption{The effects of $\lambda$ (left), $\delta$ (middle), and $\mathbf{x}_0$ (right) on gradient descent.}
\end{figure}

@@

With large datasets, it might be inefficient to compute a gradient over the entire dataset for each iteration. One way to address this is to use stochastic gradient descent, which uses a pointwise gradient instead. In this experiment, we use least-squares regression, which has a pointwise gradient of 
\begin{align*}
J(\theta; x^{(i)}, y^{(i)}) = (x^{(i)}\theta_t - y^{(i)})^2
\end{align*}
On each iteration, we update our value as follows:
\begin{align*}
\theta_{t+1} = \theta_{t} - \eta_t \nabla_\theta J(\theta_t; x^{(i)}, y^{(i)})
\end{align*}
where $\nabla = (\tau_0 + t)^{-\kappa}$ is the \emph{learning rate}, and $\tau_0$ and $\kappa$ are constants.

@@

When the objective function doesn't have a closed form solution, one can approximate the gradient using a \emph{central finite difference}. In other words, we have 
\begin{align*}
\nabla f(x) \approx \nabla_h[f](x) = \dfrac{f(x+h/2) - f(x-h/2)}{h}
\end{align*}
Therefore, we only need the objective function to approximate the gradient. Note that as $h$ approaches zero, this approaches the actual value of the gradient; and as $h$ increaes, the value diverges. Figure \ref{fig:cfd} illustrates the impact of $h$ on the accuracy of the gradient: even for large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.

@@

One approach to this problem is to model $f$ as the linear combination of \emph{basis functions}. In other words, we manually choose functions $\phi_1(x), \dots, \phi_n(x)$ and algorithmically pick coefficients $c_1, \dots, c_n$, which we put together as
\begin{align*}
f(x) &= c_1 \phi_1(x) + \dots + c_n \phi_n(x).
\end{align*}
We can rewrite this with 
\begin{align*}
\bm{c} &= \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix} \\
\bm{\phi}(x) &= \begin{bmatrix} \phi_1(x) & \cdots & \phi_n(x) \end{bmatrix}
\end{align*}
so that 
\begin{align}
f(x) &= \bm{c} \cdot \bm{\phi}(x). \label{eq:predictor}
\end{align}

@@

To counteract this effect, the learning rate for the stochastic gradient descent needs to be as low as possible for. In this case, we chose $\eta_t = (r + 0.1)^{-0.51}$; with parameters any larger, the result has very little correlation with the actual function. This is likely because a large learning rate would cause smaller-degree terms to have an relatively large weight during the first few iterations, which would overtake the higher-degree terms. Having a smaller, slower-growing learning rate lets the higher-degree terms affect the result for more iterations, and therefore the result more closely fits the input data.

@@

To illustrate in an example, we have three sets of data: set A, set B, and a validation set. We perform the above process twice, once training on A and testing on B, and once vice versa. In both cases we try a variety of values of $\lambda$ and $n$ then evaluate the SSE on the validation set. The results are given in Figure \ref{fig:validationtable}
\begin{figure}[ht!]
  \centering
  \label{fig:validationtable}
  \begin{tabular}{c|cccccc}
    & 1e-05 & 0.0001 & 0.001 & 0.01 & 0.1 \\ \hline
  0 & 62.00 & 62.00 & 62.00 & 62.01 & 62.06 \\
  1 & 2.90 & 2.90 & 2.90 & 2.90 & 2.94 \\
  2 & 2.35 & \textbf{2.35} & 2.35 & 2.36 & 2.43 \\
  3 & 2.37 & 2.37 & 2.37 & 2.39 & 2.55 \\
  4 & 3.58 & 3.58 & 3.58 & 3.59 & 3.67 \\
  \end{tabular}
  \qquad
  \begin{tabular}{c|ccccc}
    & 0 & 0.3 & 0.6 & 0.9 & 1.2 \\ \hline
  0 & 72.44 & 70.95 & 69.63 & 68.47 & 67.44 \\
  1 & 35.19 & 34.09 & 33.14 & 32.34 & 31.67 \\
  2 & 38.53 & 38.65 & 38.76 & 38.87 & 38.96 \\
  3 & 30.30 & 28.46 & \textbf{28.09} & 28.47 & 29.21 \\
  4 & 130.20 & 67.54 & 51.63 & 46.17 & 44.35 \\
  \end{tabular}
  \caption{SSE values on validation set after training with various $n$ (rows) and $\lambda$ (columns). Left: training on set A. Right: training on set B. Best values bolded.}
\end{figure}
The SSE for the first case with $n = 2$ and $\lambda = 0.0001$, tested on set B, is $25.752$. In the second case, tested on set A with $n = 3$ and $\lambda = 0.6$, the SSE is $40.087483697$. Testing with parameters $n=4$ and $\lambda = 0.1$ for both cases, which the table would predict to be suboptimal, indeed gives bigger test SSEs of 31.1388659319 and 64.9392925156, respectively.

@@

LASSO extends linear regression as well, but in a slightly different way. The balance between minimizing error and minimizing coefficient size is achieved by finding the optimal $\bm{c}$ for the objective function
\begin{align}
|\bm{y} - X \bm{c}|^2 + \lambda \sum_{i=1}^{n} |c_i|.
\end{align}
In other words, we use the sum of the first order magnitudes of the coefficients rather than second order.

This gives a similar benefit as ridge regression in that it can be tuned to generalize, avoiding overfitting and underfitting. However, it additionally has a subtle but important advantage: LASSO encourages coefficients to be exactly zero, resulting in sparse coefficient vectors. This can be useful when the basis dimension is large by simplifying and speeding up computations (as all zero coefficients can be ignored) and also allows for the strategy of throwing all possible basis functions at the regression and letting LASSO find which functions to use.

To intuitively see why LASSO encourages exactly zero coefficients, roughly consider applying gradient descent. If at a certain point the gradient in one variable is larger than that of another, it would make the objective function (specifically sum of absolute values) smaller to transfer the combined sum of the two variables all to only one of the variables, making the other one zero.

\subsection{Sparsity in LASSO versus Ridge Regression}

We illustrate the sparsity property of LASSO in an example where all the data points were generated using a sparse set of coefficients on a given set of basis functions plus some noise. The data points are divided into training, validation, and test sets. We train prediction curves with the same basis functions using linear regression (when $\lambda = 0$), ridge regression (parameters chosen using validation), and LASSO (parameters chosen using validation).

Inspecting the coefficient vectors numerically, we find that linear and ridge regression have all nonzero coefficients, but LASSO gives nonzero coefficients exactly for the same functions as the true generating coefficients. Indeed, in this case LASSO correctly identified the correct basis functions. This is the ``selection operator'' that makes up the last two letters of LASSO.

If we plot all the data points, prediction curves, and generating (true) curve in Figure \ref{fig:everything}, we see that all regressions give a relatively good prediction of the data. All the methods have generally find coefficients that give a close approximation of the training data, and in this case don't overfit. Depending on the application, sparsity may or may not be desired, and the method for solving the prediction problem can be appropriately chosen with consideration of each method's unique properties.

\begin{figure}[H]
  \centering
  \label{fig:everything}
  \includegraphics[width=0.5\textwidth]{../images/everything.png}
  \caption{Comparison of various prediction curves plotted with data points.}
\end{figure}
